Curvature approximation on surfaces -- A Discrete Exterior Calculus Approach

### 2014-11-27 article/main.tex
...

  \subsection{Geometrical operators}
    The two momentous operators on chains are the boundary operator \( \partial \) and the star operator \( \star \).
    These are the geometrical tools for the discrete exterior derivation \( \d \) resp. the discrete Hodge star \( * \),
    what we will see below.

@@

  \subsection{DEC operators}
    For the definition of a discrete version of the Hodge star \( * \) or the exterior derivative \( \d \) we can fall back to the
    geometric operators.

    \subsubsection{The discrete Hodge star operator}

>>>>

  \subsection{Geometrical operators}
    The two momentous operators on chains are the boundary operator \( \partial \) and the star operator \( \star \).
    These are the geometrical tools for the discrete exterior derivation \( \exd \) resp. the discrete Hodge star \( * \),
    what we will see below.

@@

  \subsection{DEC operators}
    For the definition of a discrete version of the Hodge star \( * \) or the exterior derivative \( \exd \) we can fall back to the
    geometric operators.

    \subsubsection{The discrete Hodge star operator}
      The discrete Hodge star \( *:\Omega_{d}^{p}(K)\rightarrow\Omega_{d}^{2-p}(*K) \) 
      is nothing more than scale the evaluation of a discrete\mh{p}{form} by the
      ratio of the dual (Voronoi) and the primal volumes.  
      \begin{align}
      \label{eqDiscreteHodge}
        \left\langle *\alpha, \star\sigma^{p} \right\rangle 
                                        &:= \frac{|\star\sigma^{p}|}{|\sigma^{p}|} \left\langle \alpha, \sigma^{p} \right\rangle
      \end{align}
      for a simplex \( \sigma^{p}\in K \) and a \( \alpha\in\Omega_{d}^{p}(K) \).
      Note that the intrinsic volume of a \mh{0}{simplex} is 1.
      (In \cite{diploma} we can see that \eqref{eqDiscreteHodge} is consistent with \( \mathcal{O}(h^{3-p}) \) for the maximum circumscribed diameter \( h \).)
      The discrete Hodge star on a discrete dual form \( \hat{\alpha}\in\Omega_{d}^{p}(*K) \) can be  obtained implicitly by the rule
      \( **\hat{\alpha} = (-1)^{p}\hat{\alpha} \), wich holds on a even-dimensional simplicial complex as well as for smooth forms on a
      even-dimensional manifold.

    \subsubsection{The discrete exterior derivative}
      The main advantage in the definition of discrete differential forms as integral evaluation of smooth forms on (abstract) simplices is
      that we are able to use the Stokes theorem \eqref{eqStokes} to describe a discrete exterior derivative
      \( \exd : \Omega^{p}_{d}(K) \rightarrow \Omega^{p+1}_{d}(K) \):
      \begin{align}
        \left\langle \exd\psi\left( \alpha \right), \sigma^{p} \right\rangle \label{eqDisExD}
                :&= \left\langle \psi\left( \exd\alpha \right), \sigma^{p} \right\rangle \\
                &= \left\langle \psi\left( \alpha \right), \partial\sigma^{p} \right\rangle \label{eqStokes}
      \end{align}
      (this applies on \( \star K \) also) and there is no approximation error accept the interpretation of the integral expression.
      Hence, the discrete exterior derivative inherit the complex property of the boundary operator, i.e. \( \exd\circ\exd=0 \).
      That's why the sequence of discrete differential forms (cochains) \( \left( \Omega^{p}_{d}(K), \exd \right)\) 
      (and  \( \left( \Omega^{p}_{d}(\star K), \exd \right) \))
      \begin{align}
      \begin{xy}
        \xymatrix{
          0 \ar[r] & 
          \Omega^{0}_{d}(K) \ar[r]^{\exd} \ar[d]^{\star} & 
          \Omega^{1}_{d}(K) \ar[r]^{\exd} \ar[d]^{\star} & 
          \Omega^{2}_{d}(K) \ar[r] \ar[d]^{\star} &
          0 \\
          0  & 
          \Omega^{2}_{d}(\star K) \ar[l] \ar[u]& 
          \Omega^{1}_{d}(\star K) \ar[l]^{\exd} \ar[u] & 
          \Omega^{0}_{d}(\star K) \ar[l]^{\exd} \ar[u] &
          0 \ar[l]
        }
      \end{xy}
      \end{align}
      is a cochain complex.

    \subsubsection{The discrete Laplace-Beltrami operator}
      The Laplace-Beltrami operator \( \Delta_{B}= * \exd * \exd \) is the special case of the negative Laplace-de Rham operator
      \( -\Delta_{dR} = *\exd *\exd + \exd *\exd * \) for \mh{0}{forms} on a \mh{2}{mainfold} or in general on a even-dimensional manifold.
      To discretize \( \Delta_{B} \) all of work is done above,
      because we can use the definitions \eqref{eqDiscreteHodge} and \eqref{eqDisExD} step by step in the same manner like the smooth exterior
      calculus.
      I.e. for a \mh{0}{form} \( f \) on a vertex \( v \) it results in
      \begin{align}
        \label{eqDisLaplace}
        \left\langle \Delta_{B} f, v \right\rangle
                &= \frac{1}{\left| \star v \right|} \sum_{\sigma^{1}=\left[ v,v_{i} \right]\in K} \frac{\left| \star\sigma^{1} \right|}{\left| \sigma^{1} \right|}
                       \left( f(v_{i}) - f(v) \right) \formpunkt
      \end{align}
      For greater details see \cite{hirani} or \cite{desbrun}. 
      \eqref{eqDisLaplace}, equivalent to the cotan-formular described in \cite{meyer}, is a approximation of order 2 (see \cite{xu})
      and was already earlier published in \cite{arakawa} as a non-DEC consequence.

      \paragraph{Note on Implementation}
        In section \ref{secCurvatureVector} we need the discrete Laplace-Beltrami operator to approximate the mean curvature.
        One way to do that is to represent the discrete linear operators \( * \) and \( \exd \) as matrices act on
        a vector of all vertices, edges or triangles.
        So the discrete Laplace-Beltrami is nothing more than the product of these matrices 
        (see \cite{siggraphKap8}, \cite{pydec} or \cite{crane}).

        A more ``FEM-like'' manner is to decompose \eqref{eqDisLaplace} in parts, which can be computed on all (triangle) elements without
        knowledge of their neighbours.
        For this we formulate the operator on the Voronoi cells to cancel the scale factor \( \left| \star v \right|^{-1} \) 
        in \eqref{eqDisLaplace}, split the sum and use a local vertex indexing on all elements:
        \begin{align} \label{eqVoronoiLaplace}
                \left\langle *\Delta_{B} f , \star v_{i} \right\rangle
                               &= \sum_{\sigma^{1}=\left[ v_{i}, v_{j} \right]} 
                     \frac{\left| \star\sigma^{1} \right|}{\left| \sigma^{1} \right|}
                      \left( f(v_{j}) - f(v_{i}) \right)\\
                               &= \sum_{\substack{\sigma^{2}=\left[ v^{\sigma^{2}}_{0},v^{\sigma^{2}}_{1},v^{\sigma^{2}}_{2} \right]
                              \\
                                               v^{\sigma^{2}}_{0} = v_{i}}}
                  \sum_{l=1,2} C^{\sigma^{2}}_{0,l} 
                          \left( f^{\sigma^{2}}_{l} -  f^{\sigma^{2}}_{0}\right) \formpunkt \label{eqLokalLaplace}
        \end{align}
        \( f^{\sigma^{2}}_{l}\in\R \) are the evaluations of \( f \) on the local vertices \( v^{\sigma^{2}}_{l} \) on the element \(
        \sigma^{2}\).
        Figure \ref{figLocalIndexing} clarify this setting.
        The coefficients \( C^{\sigma^{2}}_{0,l}\in\R \) are generally defined by
        \begin{align}
               C^{\sigma^{2}}_{k,l} &:= C^{\sigma^{2}}_{l,k}
                            := \frac{\left| \star \left[ v^{\sigma^{2}}_{k}, v^{\sigma^{2}}_{l} \right] 
                                            \cap \sigma^{2}\right|}
                                   {\left| \left[ v^{\sigma^{2}}_{k}, v^{\sigma^{2}}_{l} \right] \right|}\formpunkt
        \end{align}
        \begin{figure}
          \centering
          \includegraphics[width=.45\textwidth]{bilder/tikz/elementSummeKante.eps}
          \caption{Two adjacent (triangle) elements share a common edge \( \sigma^{1}=\left[ v_{i}, v_{j} \right] \).
                   The vertices are local indexed on these elements.}
          \label{figLocalIndexing}
        \end{figure}
        Hence we can extract from \eqref{eqLokalLaplace} a local coefficients matrix
        \begin{align}
          A^{\sigma^{2}} 
              &:= {\footnotesize\begin{bmatrix}
                    -\left( C^{\sigma^{2}}_{01} + C^{\sigma^{2}}_{02}\right) & C^{\sigma^{2}}_{01} & C^{\sigma^{2}}_{02} \\
                    C^{\sigma^{2}}_{01} & -\left( C^{\sigma^{2}}_{01} + C^{\sigma^{2}}_{12}\right) & C^{\sigma^{2}}_{12} \\
                    C^{\sigma^{2}}_{02} & C^{\sigma^{2}}_{12} & -\left( C^{\sigma^{2}}_{02} + C^{\sigma^{2}}_{12}\right)
                  \end{bmatrix}} 
        \end{align}
        which holds all computation parts for the local vertices.
        All \( A^{\sigma^{2}} \) can be assembly in a familiar way to get a sparse global matrix 
        \( A\in\R^{N_{\sigma^{0}} \times N_{\sigma^{0}}} \), where \( N_{\sigma^{0}} \) are the number of vertices in \( K \).
        At the end, for \( f_{h}\in\R^{N_{\sigma^{0}}} \), the vector of evaluations for \( f \) 
        (resp. vector of DOFs\footnote{Degree Of Freedom}), 
        \( Af_{h} \) is the result of \eqref{eqVoronoiLaplace} on all global vertices.

        The great advantage of this proceeding is the compatibility to existing Finite Element code 
        like AMDiS\footnote{Adaptive MultiDimensional Simulation toolbox, developed at Institute for Scientific Computing, TU-Dresden}
        \cite{amdis}.
        Only the element operator \( A^{\sigma^{2}} \) must be implemented 
        under use of linear test functions for the interpolation. 
        All other works, like mesh management, problem formulation on the user level, assembling and solving of the linear System, 
        can be done by the FE-toolbox.

    \subsubsection{The discrete gradient}
      The gradient of a scalar valued function \( f \) can be obtained by rising the indices of its exterior derivative, 
      in other words \( \nabla f = \left( \exd f \right)^{\sharp} \).
      Unfortunately, the sharp operator \( \sharp \) depends on the (unknown) metric of the manifold. 
      In \cite{hirani} it is suggested to use this discrete sharp operator for such a exact \mh{1}{form} \( \exd f \) to get a discrete
      gradient:
      \begin{align}
        \label{eqGradientHeine}
        \left\langle \left( \exd f \right)^{\sharp} , \star\sigma^{2}\right\rangle
          &:= \sum_{\sigma^{0}\prec\sigma^{2}} \left( f(\sigma^{0}) - f(w) \right) \nabla\Phi_{\sigma^{0}}^{\sigma^{2}} \formkomma
      \end{align}
      with a arbitrary vertex \( w\prec\sigma^{2} \) and 
      \( \Phi_{\sigma^{0}}^{\sigma^{2}} \) is the linear test function (hat function) for the vertex \( \sigma^{0} \) 
      restricted to \( \sigma^{2} \).

\section{Curvature approximations}
  \subsection{Curvature vector} \label{secCurvatureVector}
  \subsection{Weingarten map} \label{secWeingarten}

\section{Results}
